Formalizing Stack Safety as a Security Property

### 2023-02-03 submission5/core.tex
some work on 1

Computer programming relies on functions (or subroutines, procedures, methods, etc.)
as a foundational abstraction. Functions are units of computation that
can call one another to realize larger computations.
%
At a low level, a function activation comprises its own data, as
well as information about its \emph{caller} function, to which it must return.
The \emph{(call) stack} is the fundamental data structure charged with realizing
this abstraction. From a security perspective, attacks on the stack are really attacks
on the fundamental functional abstraction itself, and are therefore every bit as
pervasive.

Indeed, the call stack is a perennial target for low-level attacks.
%
The most recent release of the authoritative CWE Top 25
Most Dangerous Software Weaknesses
list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
shows general classes of vulnerabilities closely related to the stack
%% \rb{(though not exclusively)}
consistently at and near the top, e.g., \#1 (CWE-787, Out-of-bounds Write), \#5 (CWE-125, Out-of-bounds Read) or \#7 (CWE-416, Use After Free).
%
Moreover, many other types of faults involve some kind of control or data flow
using data stored in the stack \cite{DBLP:conf/raid/VeendCB12,
  DBLP:conf/sp/SzekeresPWS13,
  DBLP:conf/sp/HuSACSL16}.\footnote{\url{https://github.com/Microsoft/MSRC-Security-Research/blob/master/presentations/2019_02_BlueHatIL/}}\footnote{\url{https://www.chromium.org/Home/chromium-security/memory-safety/}}

The variety in attacks on the stack is mirrored by the
software and hardware protections that aim to prevent them,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges22:TempsDesCerises},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}.

Of these mechanisms, many are fundamentally ill-suited for offering formal guarantees,
as they aim to impede attackers but cannot provide universal protection. Shadow stacks,
for instance, aim to ``restrict the flexibility available in creating gadget chains''
\cite{Shanbhogue+19}, not to categorically
rule out attacks. Other mechanisms are stronger, but formal
guarantees are rarely discussed.
Indeed, the Cerise work of Skorstengaard et al. \cite{SkorstengaardSTKJFP} and
Georges et al. \cite{Georges22:TempsDesCerises} is so far the sole line of work
to offer (various)\bcp{Maybe we can omit that word here and spell it
  out in the related work.} stack safety formalizations.

Meanwhile, Roessler and DeHon \cite{DBLP:conf/sp/RoesslerD18} present an array of tag-based
micro-policies for stack safety, at least some of which should be amenable to making
formal guarantees. Among these is their ``Lazy Tagging and Clearing'' \(LTC\) micropolicy,
which is is peculiar in that function activations may write into one another's stack frames,
as long as the data cannot be accessed afterward. As a result, \(LTC\) definitely would not
fulfill a tag-based adaptation Georges et al.'s state-of-the-art property.

We believe that, nevertheless, \(LTC\) enforces stack safety.
It does not always protect individual addresses on the stack, but it aims
to prevent one function activation from interfering with another's behavior---to protect
the functional abstraction. We argue that a formal characterization of \(LTC\)'s protections
should fall under the umbrella of stack safety.

In this work, we propose such a characterization, built from first principles
in the hopes that it will be broadly applicable, and demonstrate
that it can distinguish between correct and flawed versions of Roessler and DeHon's
micro-policies. We use the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into a family of
properties describing
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control-flow protection
of {\em well-bracketed control flow}~\cite{SkorstengaardSTKJFP} (\(\wbcf\)).

The structure of our properties enables them to be easily extended to systems with
sophisticated features such as callee-save registers, argument-passing on the stack,
and tailcall-elimination. We likewise extend \(LTC\) to protect these features, and
test the extended micro-policy. We find that, apart from the extensions, \(LTC\)
is flawed in a way that undermines integrity and confidentiality, and we correct
this flaw.

\paragraph{Towards a Formal Property for Stack Safety}

@@

For instance, the current state-of-the-art stack-safety specification from Georges et al.
\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
control flow, and therefore defines {\em Well-bracketed Control Flow} (WBCF) under the assumption that
a callee should always return to its caller. But, in the presence of tail-call elimination, a
return from a callee to a distant ancestor is perfectly reasonable. Related, their predecessor
Skorstengaard et al. \cite{SkorstengaardSTKJFP} made no mention of confidentiality, even though
their enforcement mechanism did protect the confidentiality of the caller (not the callee.)

@@

In particular, many enforcement techniques focus purely on
well-bracketed control flow. Stack canaries aim to prevent certain attacks on the return
address, and shadow stacks with protection (e.g. Return Address Defender \cite{Chiueh2001RAD})
to enforce it completely. Others combine this protection with some degree of memory protection,
chiefly focusing on integrity. Interestingly, Skorstengaard et al. \cite{SkorstengaardSTKJFP}
describe their ``local state encapsulation'' in terms of integrity, but it is equivalent
to the combination of \(\clri\) and \(\clrc\). Their state-of-the-art successor,
Georges et al. \cite{Georges22:TempsDesCerises}, cite an early preprint of our work
as inspiration for extending their formal discussion to include confidentiality.
When checking if a preliminary version of these properties applied to
the Cerise calling convention, they noted that the Skostengaard et al. calling convention
did not enforce callee confidentiality, and made sure that their version would
\cite{Georges22:personalcommunication}, in addition to building it into their formalism.
This demonstrates the benefit of our choice to explicitly state properties in security
terms: specifying security mechanisms is hard, and when the specification takes the
form of a ``correct by construction'' machine, it is easy to neglect a non-obvious
security requirement.

@@

We now consider how {\tt f} might misbehave and violate desirable
stack safety properties associated with {\tt main}. To put the violations in a
security framework, suppose that {\tt f} is actually an attacker seeking
to leak {\tt secret}. It might do so in a number of ways, shown as snippets of
assembly code in Figure \ref{fig:f}.
%
Leakage is most obviously viewed as a violation of {\tt main}'s {\it confidentiality}.
In Figure \ref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in Figure \ref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(Figure \ref{subfig:integrity}).
Attacks of this kind do not violate {\tt main}'s confidentiality, but
rather its {\it integrity}.
In Figure \ref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (WBCF.)
%
In Figure \ref{subfig:WBCF2}, a different attack violates WBCF, this time
by returning to the correct program counter but with the wrong stack pointer.
\footnote{We pad the last two variants with {\tt nop}s just so that all the
snippets have the same length, which simplifies some later explanations.}

@@

Before the call, the program counter is 16 and the stack pointer is 980.
So we define a predicate on states that should hold just after the return:
\(\ret\ \mach \triangleq \mach[\PCname] = 20 \wedge \mach[\SP] = 980\).
%
We can identify the point just after the return (if a return occurs)
as the first state in which the pending call stack is smaller than it was
just after the call.
WBCF requires that if \(\mach\) is the state at that point, then \(\ret ~ \mach\) holds.
%For nested calls, where the pending stack is initially larger, the same principle
%applies: \(\ret ~ \mach\) must hold the next time the pending stack is the same size or smaller.

@@

Like WBCF, stack integrity defines a condition at the call that must hold upon
return. This time the condition applies to all of the memory in the function's
frame. In Figure \ref{fig:exec1} we see the lifecycle of an allocated frame:
upon allocation, the view labels it \(\object\), and when a call is made, it instead
becomes \(\sealed\). Intuitively, the integrity of {\tt main}
is preserved if, when control returns to it, any \(\sealed\) elements
are identical to when it made the call.
%
Again, we need to know when a caller has been returned to,
and we use the same mechanism of checking the depth of the call stack.
%
In the case of the call from {\tt main} to {\tt f}, the \(\sealed\) elements are the
addresses 980 through 999 and callee-saved registers such as
the stack pointer. Note that callee-saved registers often change
during the call---but if the caller accesses them after the call, it should find them
restored to their prior value.

@@

In our example setting, the observation trace consists of the sequence
of values written to {\tt out}.
The example in Figure \ref{subfig:integrity} modifies the value of {\tt sensitive},
which is \(\sealed\). Figure \ref{fig:variant} shows the state just after the call at step 5,
assuming that {\tt sec} is 5. Similar to WBCF, we define
\(\intProp\) as a predicate on states that holds if
all relevant sealed addresses in \(\mach\) are the same as after step 5.
We require that \(\intProp\) hold on the state following the matching return,
which is reached by step 9. Here {\tt sensitive} has obviously changed, but is it relevant?
Consider a variant state in which {\tt sensitive} has any other value, arbitrarily
choosing 43. As execution continues after the return
from the original state, it passes the {\tt if}-test on {\tt sensitive}, whereas the execution
from the variant does not, resulting in differing outputs. Therefore {\tt sensitive} \emph{is} relevant,
so \(\intProp\) does not hold, and integrity has indeed been violated.

>>>>

Computer programming relies on functions (or subroutines, procedures, methods, etc.)
as a foundational abstraction. Functions are units of computation that
can call one another to realize larger computations in a modular fashion.
%
At a low level, a function activation manages its own data, such as local variables, as
well as information about its \emph{caller} function, to which it must return.
The \emph{(call) stack} is the fundamental data structure charged with realizing
this abstraction in memory. \an{registers?} From a security perspective, attacks on the stack are really attacks
on the fundamental functional abstraction itself. %, and are therefore every bit as
%pervasive.

Indeed, the call stack is a ancient~\cite{one06phrack} and perennial~\cite{mitre-cwe,DBLP:conf/raid/VeendCB12,
  DBLP:conf/sp/SzekeresPWS13,
  DBLP:conf/sp/HuSACSL16,msrc-bluehat,chromium-security}
  target for low-level attacks. \apt{characterize the variety of attacks?}
%
%The most recent release of the authoritative CWE Top 25
%Most Dangerous Software Weaknesses~\cite{
%list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
%shows that general classes of vulnerabilities closely related to the stack
%% \rb{(though not exclusively)}
%consistently rank at and near the top of the list, e.g., \#1 Out-of-bounds Write, \#5 Out-of-bounds Read, and \#7 Use After Free.
%
%Moreover, many other types of faults involve some kind of control or data flow
%using data stored in the stack 
The variety in attacks on the stack is mirrored by the
software and hardware protections that aim to prevent them,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges22:TempsDesCerises},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18,Gollapudi+23}.

\apt{Add brief case for having formal guarantees.}
Of these mechanisms, many are fundamentally ill-suited for offering formal guarantees,
as they aim to impede attackers but cannot provide universal protection. Shadow stacks,
for instance, aim to ``restrict the flexibility available in creating gadget chains''
\cite{Shanbhogue+19}, not to categorically
rule out attacks. Other mechanisms are stronger, but formal
guarantees are rarely discussed.
Indeed, the Cerise work of Skorstengaard et al. \cite{SkorstengaardSTKJFP}\apt{that's not Cerise} and
Georges et al. \cite{Georges22:TempsDesCerises} is so far the sole line of work
to offer (various)\bcp{Maybe we can omit that word here and spell it
  out in the related work.} stack safety formalizations.

\apt{Transition}
Roessler and DeHon \cite{DBLP:conf/sp/RoesslerD18} present an array of tag-based
micro-policies~\cite{pump_oakland2015} for stack safety, at least some of which should be amenable to making
formal guarantees. Among these is their ``Lazy Tagging and Clearing'' (\(LTC\)) micro-policy,
which is is peculiar in that function activations may write into one another's stack frames,
as long as the data cannot be accessed afterward. Roessler and DeHon themselves say that
\(LTC\) does not enforce memory safety, but rather a form of data-flow integrity~\cite{castro+06}.
And \(LTC\) definitely would not enforce a tag-based adaptation of Georges et al.'s property.

Neverthess, we believe that there is a useful sense in which \(LTC\) does (attempt to) enforce stack safety. 
It does not always protect individual addresses on the stack, but it tries to 
prevent one function activation from interfering with another's behavior---protecting 
the functional abstraction, which we argue is the real essence of stack safety.
%We argue that a formal characterization of \(LTC\)'s protections
%should fall under the umbrella of stack safety.

In this work, we propose a formal characterization of stack safety based on the idea
of protecting each function activation from other activations.
We use the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into a family of
properties describing
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control-flow protection
of {\em well-bracketed control flow} (\(\wbcf\))~\cite{SkorstengaardSTKJFP}.
While our properties are motivated by a desire to specify \(LTC\), 
they are built from first principles in the hopes that they will be broadly applicable
to a range of enforcement mechanisms.

The structure of our properties enables them to be easily extended to systems with
sophisticated features such as callee-save registers, argument-passing on the stack,
and tailcall-elimination. We likewise extend \(LTC\) to protect these features, and
test the extended micro-policy against our property. We find that, apart from the extensions, \(LTC\)
is flawed in a way that undermines integrity and confidentiality, and we correct
this flaw.

\paragraph*{Towards Formal Properties for Stack Safety}

@@

For instance, the stack-safety specification from Georges et al.
\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
control flow, and therefore defines {\em Well-bracketed Control Flow} (\(\wbcf\)) under the assumption that
a callee should always return to its caller. But, in the presence of tail-call elimination, a
return from a callee to a distant ancestor is perfectly reasonable. Similarly, the preceding work by
Skorstengaard et al. \cite{SkorstengaardSTKJFP} made no mention of confidentiality, even though
their enforcement mechanism did protect the confidentiality of the caller (though not the callee).

@@

In particular, many enforcement techniques focus purely on
well-bracketed control flow. Stack canaries aim to prevent certain attacks on the return
address, and shadow stacks with protection (e.g. Return Address Defender \cite{Chiueh2001RAD})
to enforce it completely. Others combine this protection with some degree of memory protection,
chiefly focusing on integrity. Interestingly, Skorstengaard et al. \cite{SkorstengaardSTKJFP}
describe their ``local state encapsulation'' in terms of integrity, but it is equivalent
to the combination of \(\clri\) and \(\clrc\). The follow-up work by 
Georges et al. \cite{Georges22:TempsDesCerises} cites an early preprint of our work
as inspiration for extending their formal discussion to include confidentiality.
When checking if a preliminary version of these properties applied to
the Cerise calling convention, they noted that the Skostengaard et al. calling convention
did not enforce callee confidentiality, and made sure that their version would
\cite{Georges22:personalcommunication}, in addition to building it into their formalism.
This demonstrates the benefit of our choice to explicitly state properties in security
terms: specifying security mechanisms is hard, and when the specification takes the
form of a ``correct by construction'' machine, it is easy to neglect a non-obvious
security requirement.

@@

We now consider how {\tt f} might misbehave and violate desirable
stack safety properties associated with {\tt main}. To put the violations in a
security framework, suppose that {\tt f} is actually an attacker seeking
to leak {\tt secret}. It might do so in a number of ways, shown as snippets of
assembly code in Figure \ref{fig:f}.
%
Leakage is most obviously viewed as a violation of {\tt main}'s {\it confidentiality}.
In Figure \ref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in Figure \ref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(Figure \ref{subfig:integrity}).
Attacks of this kind do not violate {\tt main}'s confidentiality, but
rather its {\it integrity}.
In Figure \ref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (\(\wbcf\).)
%
In Figure \ref{subfig:WBCF2}, a different attack violates \(\wbcf\), this time
by returning to the correct program counter but with the wrong stack pointer.
\footnote{We pad the last two variants with {\tt nop}s just so that all the
snippets have the same length, which simplifies some later explanations.}

@@

Before the call, the program counter is 16 and the stack pointer is 980.
So we define a predicate on states that should hold just after the return:
\(\ret\ \mach \triangleq \mach[\PCname] = 20 \wedge \mach[\SP] = 980\).
%
We can identify the point just after the return (if a return occurs)
as the first state in which the pending call stack is smaller than it was
just after the call.
\(\wbcf\) requires that if \(\mach\) is the state at that point, then \(\ret ~ \mach\) holds.
%For nested calls, where the pending stack is initially larger, the same principle
%applies: \(\ret ~ \mach\) must hold the next time the pending stack is the same size or smaller.

@@

Like \(\wbcf\), stack integrity defines a condition at the call that must hold upon
return. This time the condition applies to all of the memory in the function's
frame. In Figure \ref{fig:exec1} we see the lifecycle of an allocated frame:
upon allocation, the view labels it \(\object\), and when a call is made, it instead
becomes \(\sealed\). Intuitively, the integrity of {\tt main}
is preserved if, when control returns to it, any \(\sealed\) elements
are identical to when it made the call.
%
Again, we need to know when a caller has been returned to,
and we use the same mechanism of checking the depth of the call stack.
%
In the case of the call from {\tt main} to {\tt f}, the \(\sealed\) elements are the
addresses 980 through 999 and callee-saved registers such as
the stack pointer. Note that callee-saved registers often change
during the call---but if the caller accesses them after the call, it should find them
restored to their prior value.

@@

In our example setting, the observation trace consists of the sequence
of values written to {\tt out}.
The example in Figure \ref{subfig:integrity} modifies the value of {\tt sensitive},
which is \(\sealed\). Figure \ref{fig:variant} shows the state just after the call at step 5,
assuming that {\tt sec} is 5. Similar to \(\wbcf\), we define
\(\intProp\) as a predicate on states that holds if
all relevant sealed addresses in \(\mach\) are the same as after step 5.
We require that \(\intProp\) hold on the state following the matching return,
which is reached by step 9. Here {\tt sensitive} has obviously changed, but is it relevant?
Consider a variant state in which {\tt sensitive} has any other value, arbitrarily
choosing 43. As execution continues after the return
from the original state, it passes the {\tt if}-test on {\tt sensitive}, whereas the execution
from the variant does not, resulting in differing outputs. Therefore {\tt sensitive} \emph{is} relevant,
so \(\intProp\) does not hold, and integrity has indeed been violated.
